Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-07-08 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
ms edits

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel
  flammability. But less is known about the mechanisms by which litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires.
  Multiple studies have established the existence of non-additive effects in
  the flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture dynamics vary across species with higher specific
  leaf area being associated with higher maximum absorption and lower litter
  bulk denisty being associated with faster drying rates. Species that produce
  more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
  species) have faster rates of litter desiccation. These differing moisture
  dynamics result in time since wetting having strongly differing effects on
  flammability across the major litter types.}

  % show a
  % relationship between time since wetting and time to ignition and spread rate,
  % with significant interaction with genus. This indicates a genus-specific
  % behavior of flammability and time since wetting. We found non-additive
  % effects in litter mixtures for both moisture content and flammability.
  % Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
  % dry faster, produced more flammable behavior (lower time to ignition, faster
  % spread rate, taller flames). Fire behavior of mixtures depended on moisture
  % content, although mixtures without Q. kelloggii had always higher moisture
  % content than predicted by the average fire behavior of the species in the
  % mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
  % non-additive behavior and at higher moisture contents (> 200\%) they exhibit
  % negative non-additive behavior.}

% A single paragraph of about 200 words maximum. For research articles,
% abstracts should give a pertinent overview of the work. We strongly encourage
% authors to use the following style of structured abstracts, but without
% headings: (1) Background: Place the question addressed in a broad context and
% highlight the purpose of the study; (2) Methods: Describe briefly the main
% methods or treatments applied; (3) Results: Summarize the article's main
% findings; and (4) Conclusion: Indicate the main conclusions or
% interpretations. The abstract should be an objective representation of the
% article, it must not contain results which are not presented and
% substantiated in the main text and should not exaggerate the main
% conclusions.

@@

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences:
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \setcounter{section}{-1} %% Remove this when starting to work on the template.
% \section{How to Use this Template}
% The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions for authors page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX related questions please contact latex@mdpi.com.
%The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision

\section{Introduction}
% The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. And now citing a book reference \cite{ref-book}. Please use the command \citep{ref-journal} for the following MDPI journals, which use author-date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Languages, Laws, Religions, Risks, Social Sciences.

Fuel moisture content is an important, and often driving, determinant of surface fire behavior. Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn [CITE]. In particular, fine dead fuels respond very readily to environmental changes [CITE]. In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}. Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}. However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.

Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}. Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread \cite{Nelson-2011}. Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases \cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary. Moisture of litter fuels varies with climate, with seasons and with short term weather within a season. Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Viney-1991}. If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.  Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires. Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.

@@

Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species.  Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. In most forests, leaf litter will be a mixture representative of the species in the stand. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. Blauw et al.\cite{Blauw+Wensink+etal-2015} suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.

% I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content. I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.

We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California. We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California. Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced time to ignition and flame spread rate. The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects. We compared leaf litter traits with the measured moisture absorption and desorption rates. Finally, we tested if the non-additive effects on flammability reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig7_consume_time.pdf}
\caption{Percent fuel consumed by time since wetting for litter of eight
  species. Curves shown indicate best fit beta regression lines. Model ANOVA
  table in Table \ref{tab:consume_time_anova} and coefficients in SI Table
  4. Horizontal jitter (up to 2 hr) was added to the plot for visualization and
  to avoid overplotting. All collections were at the times labeled on the
  x-axis.. A lack of data at earlier time points for a taxon indicates no
  ignitions occurred at those time points.}
  \label{fig:spread_time}
\end{figure}

@@

@@

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig9_mixture_obs_vs_pred_mc.pdf}
\caption{Observed vs predicted fuel moisture content for litter mixtures. Predictions are based on the mean moisture content at each particualr time sinece wetting for the three constituent species in each mixture.  Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_mc}
\end{figure}

@@

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
\caption{Observed vs predicted flame spread rate for litter mixtures. Y axis shows relative spread rate values (predicted - observed) so that positive values indicate faster than predicted and negative slower than predicted. Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_spread}
\end{figure}

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
\caption{Observed vs predicted fuel consumption for litter mixtures. Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_consume}
\end{figure}

@@

% In general, the flammability of mixtures was higher as time since wetting went up as expected (Table 2.2). The absence of Q. kelloggii in a mixture, in general, lead to slower flame spread rates, longer flaming combustion, and longer ignition times. The combination of both Q. kelloggii and P. jeffreyi in a mixture generated taller flames and shorter flaming combustion.

% Non-additivity in moisture content

% We examined the relationship between observed moisture content of a mixture, and the moisture content predicted based on the constituent species to determine the existence of non-additive effects in moisture content (Fig. 2.6). The mixture that does not contain Q. kelloggii (AbCaPi) consistently retains more moisture that what would be predicted by the average of its constituent species. In general, mixtures start the drying process drier than expected and after 48 hours are wetter than expected, indicating lower moisture absorption and slower desiccation rates. The non-additive behavior is dependent on the moisture content of the mixture. When moisture content is above 200\%, I observe negative non-additivity for all mixtures containing Q. kelloggii. There is additive behavior when moisture content is between 75 and 150\%, and positive non-additivity below 75\% moisture content.

% Non-additivity in flammability

% We determined the presence of non-additivity in the flammability of litter mixtures by comparing the flammability of the mixtures to a null model created by the arithmetic average of the predicted individual flammability of each species in the mixture (Figs. 2.7 and 2.8). Both spread rate and time to ignition exhibited non-additivity, mediated by time since wetting. Observed spread rate values for the mixtures were higher than predicted by the null model, though the non-additivity was stronger for mixtures that contained Q. kelloggii. This result is likely driven by the bulk density of the mixtures, which was lower than predicted, making the mixtures’ litter beds more aerated. Non-additivity in time to ignition is strongly influenced by the presence of C. decurrens in the mixture. This species failed to ignite several times in the individual trials and was assigned the maximum number of seconds for the ignition attempt, 180.

%\begin{listing}[H]
%\caption{Title of the listing}
%\rule{\textwidth}{1pt}
%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
%\rule{\textwidth}{1pt}
%\end{listing}

@@

\subsection{Leaf traits influence on moisture dynamics }

The species in this study displayed different patterns of both moisture absorption and desorption (with Q. kelloggii and the pines exhibiting higher values for both), which were controlled by different litter traits (leaf thinness and litter bulk density, respectively). Q. kelloggii was the species that absorbed the most moisture, followed by the pines. The species that have a greater absorptive capacity are also the species that are usually more flammable when dry. This indicates that following a rain event, the most flammable species in my study would not be available as fuel immediately, but cannot be discounted as potential fuel. A species capacity to absorb moisture is controlled by leaf particle geometry, specifically the ratio between leaf length and leaf thickness, and not litter bed arrangement. The eight species differed in their desorption capacity. Q. kelloggii and the pines had higher desiccation rates, while C. decurrens exhibited lower desiccation rates. The ANOVA results indicate that leaf thinness best explains variation in maximum water retention, while litter bulk density best explains variation in desiccation rates.

NOTE: P lambertiana higher water absorption, presumably due to higher SLA.

Bulk densities measured here closely match those for these and other species \cite{van_Wagtendonk+Sydoriak+etal-1998,Stephens+Finney+etal-2004}

The overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density). Decomposition can also affect moisture content in litter. Van Wagner (1969), investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.

TODO

With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density. Litter density is largely driven by leaf size in this system \cite{Magalhaes+Schwilk-2012}.

\subsection{Time since wetting and flammability}

@@

Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content. Q. kelloggii and the pines are more flammable at higher moisture contents when compared to volatile-rich species such as C. decurrens, Abies spp. or S. giganteum. This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents. I have determined moisture of extinction for the eight species in this study. As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp. and S. giganteum). I used a higher cut-off moisture for the burn trials than the moisture of extinction found in the literature (65\% vs. 27-30\%; \citep{ Rothermel-1972} [TODO Dickinson et al. 2016] in order to be conservative, as moisture content studies as sparse. However, Q. kelloggii was still flammable at ~80-90\% moisture content (values at 72 hours used for the species mixtures), which I did not anticipate. Due to its high moisture content, Q. kelloggii was not burnt at 24 and 48 hours since wetting, which explains the gap shown in the Figures. Though I cannot comment on the exact behavior of Q. kelloggii at 24 and 48 hours, it is unlikely to be more flammable than the pines. This result indicates that though Q. kelloggii might be too wet to start a fire right after a rain event, Q. kelloggii litter dries faster than the other species in this study and will become very flammable within two days.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires.
  Multiple studies have established the existence of non-additive effects in
  the flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture dynamics vary across species with higher specific
  leaf area being associated with higher maximum absorption and lower litter
  bulk denisty being associated with faster drying rates. Species that produce
  more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
  species) have faster rates of litter desiccation. We examined two axes of
  flammability (spread rate and fuel consumption) and found that these
  differing moisture dynamics result in time since wetting having strongly
  differing effects on flammability across the major litter types typified by
  different tree genera. Furthermore, we found that litter mixtures exhibited
  consistent positive non additivity in flame spread rate and that fuel
  consumption and flame spread rate both showed increasing positive non
  additivity with time since wetting.

@@

@@

\section{Introduction}

Fuel moisture content is an important, and often driving, determinant of surface fire behavior. Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn \cite{Kreye+Varner+Etal-2018. In particular, fine dead fuels respond very readily to environmental changes \cite{Nelson 2001}.  Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}. Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread \cite{Nelson-2011}. Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases \cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary. Moisture of litter fuels varies with climate, with seasons and with short term weather within a season. Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Nelson 2001}.

Fire scientists have long categorized fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}. A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite[eg]{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001} that include at least temperature and relative humidity and most aim to describe equilibrium conditions although some include precipitation and solar radiation effects as well. Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and therefore fire behavior can depend on rate of drying following a wetting event. If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting. Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires.

In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}. Similarly, strong species specific effects on litter flammability are recorded for the forests tof the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a strong role in flammability feedbacks \cite{Kreye+Varner+etal-2013}.  Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}. However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.

@@

Flammability is a multi-dimensional characteristic of fuels \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing different components of fire behavior. In litter driven fire, the two major axes of variation are described by measurements related to flame spread rate and measurements related to heat transfer. Flame spread rate is easy to measure in burning trials, but heat transfer can be more difficult and often proxies such as duration of heating are used \citep{Magalhaes+Schwilk-2012, Varner+Kane+etal-2015}. These different components of fire behavior have differing ecological effects. Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species. Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. In most forests, leaf litter will be a mixture representative of the species in the stand. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. Blauw et al.\cite{Blauw+Wensink+etal-2015} suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.

We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California. We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California. Previous work described the flammability of these species in monoculture and in mixture for dry litter \cite{Magalhaes+Schwilk-2012}. {TODO- more on oaks?]  Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed. The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects. We compared leaf litter traits with the measured moisture absorption and desorption rates. Finally, we tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig7_consume_time.pdf}
\caption{Percent fuel consumed by time since wetting for litter of eight
  species. Curves shown indicate best fit beta regression lines. Model ANOVA
  table in Table \ref{tab:consume_time_anova} and coefficients in SI Table
  4. Horizontal jitter (up to 2 hr) was added to the plot for visualization and
  to avoid overplotting. All collections were at the times labeled on the
  x-axis.. A lack of data at earlier time points for a taxon indicates no
  ignitions occurred at those time points.}
  \label{fig:consume_time}
\end{figure}

@@

\subsection{Non additivity in  litter mixtures}

@@

\begin{figure}[H]
  \centering
\includegraphics[width=16cm]{figs_tables/fig9_mixture_obs_vs_pred_mc.pdf}
\caption{Observed vs predicted fuel moisture content for litter mixtures. Predictions are based on the mean moisture content at each particualr time sinece wetting for the three constituent species in each mixture.  Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_mc}
\end{figure}

@@

\begin{figure}[H]
  \centering
\includegraphics[width=16cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
\caption{Observed vs predicted flame spread rate for litter mixtures. Y axis shows relative spread rate values (predicted - observed) so that positive values indicate faster than predicted and negative slower than predicted. Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_spread}
\end{figure}

\begin{figure}[H]
  \centering
\includegraphics[width=16cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
\caption{Observed vs predicted fuel consumption for litter mixtures. Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
  \label{fig:mixture_obs_pred_consume}
\end{figure}

@@

@@

The species in this study displayed different patterns of both moisture absorption and desorption, which were controlled by different litter traits (specific leaf area and litter bulk density, respectively). \emph{Q. kelloggii} was the species that absorbed the most moisture, followed distantly by the finest-needled pine, \emph{P. lambertiana}. These species that have a greater absorptive capacity are also among the most flammable in terms of spread rate when dry. One presumption of ours prior to this experiment was that the high flammability previously reported for \emph{Q. kelloggii} \cite{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture contents. This was partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours of drying in this study (Figs. \ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture content in this oak was unable to suppress ignition of mixtures to which it contributed (Figs. \ref{{fig:mixture_obs_pred_spread} and \ref{{fig:mixture_obs_pred_consume}). 
NOTE: P lambertiana higher water absorption, presumably due to higher SLA.

With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density. Litter density is largely driven by leaf size in this system \cite{Magalhaes+Schwilk-2012}.  The overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density). Decomposition can also affect moisture content in litter. Van Wagner (1969), investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.

T\subsection{Time since wetting and flammability}

@@

Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content. The pines are more flammable at higher moisture contents when compared to volatile-rich species such as \emph{C. decurrens}, \emph{Abies spp}. or \emph{S. giganteum}. This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents.

% I have determined moisture of extinction for the eight species in this study. As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp. and S. giganteum). I used a higher cut-off moisture for the burn trials than the moisture of extinction found in the literature (65\% vs. 27-30\%; \citep{ Rothermel-1972} [TODO Dickinson et al. 2016] in order to be conservative, as moisture content studies as sparse. However, Q. kelloggii was still flammable at ~80-90\% moisture content (values at 72 hours used for the species mixtures), which I did not anticipate. Due to its high moisture content, Q. kelloggii was not burnt at 24 and 48 hours since wetting, which explains the gap shown in the Figures. Though I cannot comment on the exact behavior of Q. kelloggii at 24 and 48 hours, it is unlikely to be more flammable than the pines. This result indicates that though Q. kelloggii might be too wet to start a fire right after a rain event, Q. kelloggii litter dries faster than the other species in this study and will become very flammable within two days.
